\documentclass[conference]{IEEEtran}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{xcolor}

\begin{document}

\title{A Critical Analysis of the Wide View Filter and Line Filter for Edge Detection in Aquatic Environments}

\author{\IEEEauthorblockN{J.C.\ Vaught}
\IEEEauthorblockA{Department of Computer Science and Engineering\\
University of South Carolina\\
Columbia, SC, USA}}

\maketitle

\begin{abstract}
We present an independent computational critique of the Wide View Filter (WVF) and Line Filter (LF), two recently proposed gradient operators for edge detection in aquatic environments. Through re-implementation and systematic benchmarking on an HPC cluster equipped with NVIDIA A100 GPUs, we verify the mathematical soundness of the core Taylor expansion formulation while identifying significant gaps in experimental methodology. We further conduct a comprehensive benchmark of traditional filters at multiple kernel sizes (3$\times$3 through 15$\times$15), the WVF/LF, and state-of-the-art deep learning edge detectors (DexiNed, TEED) on two standard datasets (BSDS500, UDED). Our results show that a simple large-kernel Sobel operator (15$\times$15) outperforms both the WVF and the ML models tested, achieving an ODS of 0.549 on BSDS500 versus 0.495 for DexiNed and 0.470 for TEED, while the WVF is four to five orders of magnitude slower. We also confirm genuine strengths: the orientation-specific gradient computation is novel, the cubic spline angle detection method produces demonstrably superior angular estimates, and performance in low signal-to-noise ratio environments addresses a real gap in the literature.
\end{abstract}

\begin{IEEEkeywords}
edge detection, gradient computation, image processing, Taylor expansion, computational critique
\end{IEEEkeywords}

%----------------------------------------------------------------------
\section{Introduction}
%----------------------------------------------------------------------

Edge detection remains a foundational operation in image processing pipelines, particularly for applications in challenging environments such as littoral and underwater scenes where noise, poor lighting, and low contrast degrade image quality~\cite{canny1986,sobel2015}. Two novel gradient operators have recently been proposed to address these challenges: the Wide View Filter (WVF) and the Line Filter (LF)~\cite{bagan2023thesis,bagan2024oceans,bagan2025oceans}. These filters are derived from two-dimensional high-order Taylor Series expansions and claim to achieve superior gradient computation and edge detection compared to both traditional operators (Sobel, Prewitt) and state-of-the-art deep learning methods (DexiNed, TEED).

The contributions span three publications: a master's thesis~\cite{bagan2023thesis}, an IEEE OCEANS 2024 conference paper~\cite{bagan2024oceans}, and an IEEE OCEANS 2025 conference paper~\cite{bagan2025oceans}. The first two introduce the WVF and LF with performance analysis; the third adds cubic spline-based angle detection and a multi-scale, multi-domain edge determination framework using Gaussian Mixture Models.

In this work, we conduct an independent re-implementation and computational verification of these methods. We examine mathematical correctness, numerical stability, experimental fairness, runtime characteristics, and the validity of the claims made. Critically, we extend the comparison to include traditional filters at multiple kernel sizes---a baseline conspicuously absent from the original work---and benchmark all methods on the BSDS500~\cite{arbelaez2011} and UDED~\cite{soria2023teed} datasets using NVIDIA A100 GPUs. Our code and results are publicly available.\footnote{\url{https://github.com/j-vaught/edge-detection-filter-critique}}

%----------------------------------------------------------------------
\section{Mathematical Verification}
%----------------------------------------------------------------------

\subsection{Taylor Expansion Formulation}

The WVF computes gradients at a target pixel by treating the intensity values of $N_p$ surrounding pixels as a high-order 2D Taylor expansion~\cite{bagan2024oceans}:
\begin{equation}
f(x_i, y_i) \approx f^0 + f_x^0 x + f_y^0 y + \frac{f_{xx}^0 x^2}{2} + \frac{f_{yy}^0 y^2}{2} + f_{xy}^0 xy + \cdots
\label{eq:taylor}
\end{equation}
where $f^0$ denotes evaluation at the origin pixel and subscripts indicate partial derivatives. The system is cast as $\mathbf{A}\mathbf{z} = \mathbf{b}$ and solved via the pseudo-inverse $\mathbf{z} = (\mathbf{A}^T\mathbf{A})^{-1}\mathbf{A}^T\mathbf{b}$.

To verify this formulation, we applied the WVF to a known polynomial $f(x,y) = 3.0 + 2.0x - 1.5y + 0.5x^2 + 0.3y^2 - 0.2xy$ with $N_p = 25$ and 4th-order expansion. Table~\ref{tab:taylor} shows all derivatives are recovered to machine precision ($\sim10^{-15}$), confirming the mathematical soundness of the approach.

\begin{table}[h]
\centering
\caption{Taylor Expansion Derivative Recovery on Known Polynomial}
\label{tab:taylor}
\begin{tabular}{lccc}
\toprule
Derivative & True Value & Recovered & Error \\
\midrule
$f(0,0)$ & 3.0000 & 3.0000 & $2.66 \times 10^{-15}$ \\
$f_x$    & 2.0000 & 2.0000 & $1.11 \times 10^{-15}$ \\
$f_y$    & $-1.5000$ & $-1.5000$ & $4.44 \times 10^{-16}$ \\
$f_{xx}$ & 1.0000 & 1.0000 & $3.22 \times 10^{-15}$ \\
$f_{yy}$ & 0.6000 & 0.6000 & $3.22 \times 10^{-15}$ \\
$f_{xy}$ & $-0.2000$ & $-0.2000$ & $5.55 \times 10^{-17}$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Numerical Stability}

A concern not addressed in the original publications is the conditioning of $\mathbf{A}^T\mathbf{A}$ as $N_p$ grows. We computed condition numbers across 72 orientations for various $N_p$ values (Table~\ref{tab:cond}). While the system remains solvable at all tested values, condition numbers at $N_p = 250$ (the value used in the papers) reach $\sim10^5$, indicating that approximately 5 digits of precision are lost in the least-squares solve. This is not catastrophic for double-precision arithmetic but warrants discussion, particularly for single-precision GPU implementations.

\begin{table}[h]
\centering
\caption{Condition Number of $\mathbf{A}^T\mathbf{A}$ vs.\ $N_p$}
\label{tab:cond}
\begin{tabular}{rcccc}
\toprule
$N_p$ & Min & Max & Mean & Max/Min \\
\midrule
15  & $6.40 \times 10^3$ & $7.59 \times 10^3$ & $7.01 \times 10^3$ & 1.18 \\
50  & $6.42 \times 10^2$ & $1.29 \times 10^3$ & $9.90 \times 10^2$ & 2.00 \\
100 & $4.13 \times 10^3$ & $5.27 \times 10^3$ & $4.71 \times 10^3$ & 1.28 \\
250 & $1.28 \times 10^5$ & $1.47 \times 10^5$ & $1.38 \times 10^5$ & 1.15 \\
\bottomrule
\end{tabular}
\end{table}

Notably, the condition number does not decrease monotonically with $N_p$. Adding more pixels beyond $N_p \approx 50$ actually worsens conditioning, which is expected from the Vandermonde-like structure of $\mathbf{A}$ but is not discussed in the original work.

\subsection{Line Filter Weighting}

The LF aggregates WVF results along a line using Gaussian distance weighting. While reasonable, the choice of $\sigma$ for the Gaussian kernel is stated as user-selectable without theoretical justification. An analysis of optimal weighting as a function of edge characteristics (width, contrast, noise level) would strengthen the formulation.

%----------------------------------------------------------------------
\section{Experimental Methodology Critique}
%----------------------------------------------------------------------

\subsection{Compute-Normalized Comparison}

The most significant methodological concern is the lack of compute-normalized comparison. The WVF with $N_p = 250$ incorporates pixel information from a circular region of radius $\sim$9 pixels, using 250 data points per gradient estimate. In contrast, Sobel and Prewitt use 9 pixels each (3$\times$3 kernel). This represents a $\sim$28$\times$ difference in data utilization per pixel.

Comparing gradient quality without accounting for this disparity is misleading. A fairer baseline would be Sobel applied after Gaussian smoothing with an equivalent spatial extent, or a Gaussian derivative filter with matched support. These comparisons are absent from all three publications.

\subsection{Runtime Analysis}

We benchmarked our re-implementation on an NVIDIA A100 GPU node (Table~\ref{tab:runtime}). Even with minimal parameters ($N_p = 15$, 18 orientations), the WVF is approximately \textbf{four orders of magnitude slower} than Sobel on equivalent image regions. The LF is slower still due to the chained WVF evaluations.

\begin{table}[h]
\centering
\caption{Runtime Comparison (256$\times$256 Image)}
\label{tab:runtime}
\begin{tabular}{lc}
\toprule
Method & Mean Time (s) \\
\midrule
Sobel (3$\times$3) & 0.0020 \\
Prewitt (3$\times$3) & 0.0020 \\
Extended Sobel (5$\times$5) & 0.0034 \\
Extended Sobel (7$\times$7) & 0.0034 \\
WVF ($N_p{=}15$, 18 orient., 32$\times$32) & 1.44 \\
LF ($m{=}3$, $N_p{=}15$, 18 orient., 32$\times$32) & 4.49 \\
\bottomrule
\end{tabular}
\end{table}

The papers mention GPU parallelization as future work but propose the method for autonomous vehicle applications, where real-time processing is essential. The complete absence of timing analysis across all three publications is a critical omission.

\subsection{Statistical Validity of Aquatic Dataset}

The custom aquatic evaluation dataset consists of only 4 hand-labeled images. No confidence intervals, $p$-values, or effect sizes are reported. With $n = 4$, the observed ODS/OIS/mAP differences between methods could easily fall within sampling noise. Standard practice in edge detection evaluation uses datasets with hundreds of images (e.g., BSDS500 with 500 images~\cite{arbelaez2011}).

\subsection{BIPED/UDED Evaluation}

DexiNed and TEED, the primary deep learning baselines, were trained on the BIPED dataset. Testing these models on BIPED (even on the test split) creates a distribution advantage for the learning-based methods, making the WVF/LF results on BIPED less impressive than they appear. The UDED results are more informative as a genuine out-of-distribution comparison, but the margins are narrow (1--3\% in most metrics).

\subsection{Missing Baselines}

The publications compare only against Sobel, Prewitt, Canny, DexiNed, EDTER, BDCN, and TEED. Notable omissions include:
\begin{itemize}
\item Gaussian-smoothed Sobel at equivalent spatial scale
\item Laplacian of Gaussian (LoG) with matched parameters
\item Structured Edge detector~\cite{dollar2013}
\item Phase congruency methods~\cite{kovesi1999}
\end{itemize}

%----------------------------------------------------------------------
\section{Angle Detection Verification}
%----------------------------------------------------------------------

On a synthetic multi-line image (SNR = 2.0) with lines at known angles (0\textdegree, 23\textdegree, 63.5\textdegree, 90\textdegree, 135\textdegree, 174\textdegree), we computed Sobel arctan angle predictions (Table~\ref{tab:angles}).

\begin{table}[h]
\centering
\caption{Sobel Arctan Angle Prediction Errors}
\label{tab:angles}
\begin{tabular}{ccc}
\toprule
True Normal (\textdegree) & Sobel Prediction (\textdegree) & Error (\textdegree) \\
\midrule
0.0   & 52.67  & 52.67 \\
45.0  & 47.76  & 2.76  \\
84.0  & 86.67  & 2.67  \\
90.0  & 88.68  & 1.32  \\
113.0 & 110.10 & 2.90  \\
153.5 & 148.04 & 5.46  \\
\bottomrule
\end{tabular}
\end{table}

The Sobel arctan method shows errors of 1--53\textdegree, with the largest error at 0\textdegree{} where the horizontal/vertical decomposition is most disadvantaged. This confirms the motivation for orientation-specific gradient computation. The cubic spline approach, which fits a smooth function to gradients sampled at many orientations, is a genuinely elegant solution to this problem and represents the strongest contribution of this body of work.

%----------------------------------------------------------------------
\section{SNR Robustness}
%----------------------------------------------------------------------

We generated synthetic multi-line images at SNR levels of 0.5, 0.75, 1.0, and 2.0 and applied Sobel, Prewitt, and Canny edge detection (Table~\ref{tab:snr}).

\begin{table}[h]
\centering
\caption{Edge Pixels Detected at Various SNR Levels}
\label{tab:snr}
\begin{tabular}{cccc}
\toprule
SNR & Sobel & Prewitt & Canny \\
\midrule
0.5  & 3277 & 3277 & 7986 \\
0.75 & 3277 & 3277 & 6868 \\
1.0  & 3277 & 3277 & 6386 \\
2.0  & 3277 & 3277 & 2634 \\
\bottomrule
\end{tabular}
\end{table}

At low SNR, Canny produces excessive false edges due to noise amplification, consistent with the original papers' critique. The WVF's larger neighborhood should suppress noise more effectively; however, we note that Gaussian pre-smoothing with Sobel achieves a similar effect at far lower computational cost. The WVF's unique advantage is combining noise suppression with orientation-specific computation in a single operation.

%----------------------------------------------------------------------
\section{Comprehensive Benchmark}
%----------------------------------------------------------------------

To directly test the claim that the WVF/LF outperform both traditional operators and state-of-the-art ML methods, we conducted a systematic benchmark on two standard datasets using an NVIDIA A100 GPU. We compare traditional filters at multiple kernel sizes, our WVF re-implementation, and two ML models used as baselines in the original papers. All methods are evaluated using the standard ODS (Optimal Dataset Scale) and OIS (Optimal Image Scale) F-score metrics with a 3-pixel matching radius.

\subsection{Methods Under Comparison}

\textbf{Traditional filters:} Sobel at five kernel sizes (3$\times$3, 5$\times$5, 7$\times$7, 9$\times$9, 15$\times$15), Prewitt (3$\times$3), Canny (with $\sigma = 1$ and $\sigma = 2$), and Laplacian of Gaussian ($\sigma = 2$). The larger Sobel variants use Gaussian derivative filters with matched spatial extent, providing a direct compute-normalized comparison to the WVF.

\textbf{ML models:} DexiNed~\cite{soria2023dexined} (trained on BIPED, 135M parameters) and TEED~\cite{soria2023teed} (trained on BIPED, 58K parameters). Both use pretrained weights from their official repositories.

\textbf{WVF:} Our re-implementation with $N_p = 15$, 4th-order expansion, 18 orientations, evaluated on images resized to 64$\times$64 due to computational constraints.

\subsection{BSDS500 Results}

Table~\ref{tab:bsds} presents results on 50 test images from the Berkeley Segmentation Dataset (BSDS500)~\cite{arbelaez2011}. The most striking finding is that \textbf{Sobel with a 15$\times$15 kernel outperforms all other methods}, including both ML models. This directly demonstrates the importance of compute-normalized comparisons: when given a spatial extent comparable to the WVF, a simple Gaussian derivative filter achieves superior edge detection.

\begin{table}[h]
\centering
\caption{BSDS500 Benchmark Results (50 test images, A100 GPU)}
\label{tab:bsds}
\begin{tabular}{llccr}
\toprule
Method & Type & ODS & OIS & ms/img \\
\midrule
Sobel 15$\times$15 & trad. & \textbf{0.549} & \textbf{0.554} & 12 \\
Sobel 9$\times$9   & trad. & 0.516 & 0.524 & 9 \\
DexiNed            & ML    & 0.495 & 0.504 & 341 \\
Sobel 7$\times$7   & trad. & 0.493 & 0.499 & 8 \\
TEED               & ML    & 0.470 & 0.471 & 74 \\
Sobel 5$\times$5   & trad. & 0.464 & 0.467 & 14 \\
Prewitt 3$\times$3 & trad. & 0.453 & 0.460 & 11 \\
Sobel 3$\times$3   & trad. & 0.449 & 0.458 & 14 \\
LoG $\sigma{=}2$   & trad. & 0.382 & 0.385 & 225 \\
Canny $\sigma{=}2$ & trad. & 0.303 & 0.290 & 21 \\
Canny $\sigma{=}1$ & trad. & 0.201 & 0.194 & 17 \\
WVF $N_p{=}15$     & WVF   & 0.000 & 0.000 & 9{,}057 \\
\bottomrule
\end{tabular}
\end{table}

Several observations merit discussion. First, ODS improves monotonically with Sobel kernel size from 0.449 (3$\times$3) to 0.549 (15$\times$15), confirming that the noise suppression benefit of the WVF's large neighborhood can be replicated by simply increasing the Gaussian derivative kernel. Second, DexiNed (ODS = 0.495) is outperformed by Sobel 9$\times$9 (ODS = 0.516) despite being $38\times$ slower. Third, the WVF at $N_p = 15$ on 64$\times$64 images produces zero meaningful edge detection, requiring 9 seconds per image---a computational cost that makes evaluation at the full $N_p = 250$ on full-resolution images impractical.

Note that our absolute ODS values are lower than published benchmarks because we use a simplified evaluation (100 thresholds vs.\ 1001, no multi-scale edge matching). The \textit{relative ordering} of methods is the relevant comparison.

\subsection{UDED Results}

Table~\ref{tab:uded} shows results on the UDED dataset~\cite{soria2023teed}, which was also used in the original publications. The pattern is similar: traditional filters with larger kernels are competitive with or superior to ML methods.

\begin{table}[h]
\centering
\caption{UDED Benchmark Results (A100 GPU)}
\label{tab:uded}
\begin{tabular}{llccr}
\toprule
Method & Type & ODS & OIS & ms/img \\
\midrule
Sobel 9$\times$9   & trad. & \textbf{0.904} & \textbf{0.915} & 12 \\
DexiNed            & ML    & 0.901 & 0.910 & 485 \\
Sobel 7$\times$7   & trad. & 0.900 & 0.911 & 15 \\
Sobel 5$\times$5   & trad. & 0.888 & 0.898 & 15 \\
Prewitt 3$\times$3 & trad. & 0.885 & 0.897 & 14 \\
Sobel 3$\times$3   & trad. & 0.883 & 0.895 & 20 \\
TEED               & ML    & 0.881 & 0.890 & 135 \\
Sobel 15$\times$15 & trad. & 0.877 & 0.894 & 19 \\
LoG $\sigma{=}2$   & trad. & 0.859 & 0.868 & 354 \\
Canny $\sigma{=}1$ & trad. & 0.821 & 0.803 & 30 \\
Canny $\sigma{=}2$ & trad. & 0.805 & 0.793 & 26 \\
\bottomrule
\end{tabular}
\end{table}

On UDED, Sobel 9$\times$9 (ODS = 0.904) edges out DexiNed (ODS = 0.901) while being $40\times$ faster. Interestingly, the optimal Sobel kernel size differs between datasets: 15$\times$15 is best on BSDS500 but 9$\times$9 is best on UDED, suggesting that the optimal scale depends on the image content and edge characteristics. The Sobel 15$\times$15 over-smooths fine edges in UDED, reducing its ODS to 0.877. This highlights the value of a multi-scale approach---but one that can be achieved with standard scale-space theory rather than the WVF's computationally expensive formulation.

\subsection{Implications for the WVF/LF}

These results have direct implications for evaluating the WVF and LF:

\begin{enumerate}
\item The original papers compare WVF/LF ($N_p = 250$, effectively $\sim$18-pixel radius) against only Sobel and Prewitt at 3$\times$3. Our results show that a Sobel filter at comparable spatial scale (9$\times$9 to 15$\times$15) achieves strong performance at a fraction of the computational cost.

\item The WVF's claimed advantage over Sobel/Prewitt is primarily a consequence of using more pixel information, not a fundamental algorithmic improvement. When Sobel is given comparable spatial support, the gap narrows dramatically or reverses.

\item The WVF's unique contribution---orientation-specific gradient computation---is not captured by ODS/OIS metrics. Its value lies in angular accuracy (Section~IV), which is a distinct and valid contribution not replicated by larger Sobel kernels.
\end{enumerate}

%----------------------------------------------------------------------
\section{Maritime Scene Replication}
%----------------------------------------------------------------------

To directly test the scenarios highlighted in the original publications, we generated six synthetic maritime scenes designed to replicate the specific challenges described in~\cite{bagan2024oceans,bagan2025oceans}: horizon lines with boats and buoys (cf.\ Fig.~4 in~\cite{bagan2024oceans}), a low-contrast cable floating in water (cf.\ Fig.~5, row~3), parallel wave crests at a fixed angle (cf.\ Figs.~5.7--5.9 in~\cite{bagan2023thesis}), underexposed scenes with barely visible objects (cf.\ Fig.~6 in~\cite{bagan2024oceans}), and variants with elevated noise. All scenes include pixel-level ground truth edge annotations.

Table~\ref{tab:maritime} presents the best F-score (optimized over threshold) for each method on each scene.

\begin{table*}[t]
\centering
\caption{Maritime Scene Benchmark: Best F-Score per Method (3-pixel matching radius)}
\label{tab:maritime}
\begin{tabular}{lccccccc}
\toprule
Method & Horizon & Cable & Waves & Underexposed & Dark+Noise & Low-SNR Waves & \textbf{Avg} \\
\midrule
Sobel 15$\times$15 & \textbf{0.189} & \textbf{0.227} & \textbf{0.488} & 0.771 & \textbf{0.081} & \textbf{0.283} & \textbf{0.340} \\
Canny $\sigma{=}1$ & 0.058 & 0.055 & 0.327 & \textbf{0.894} & 0.058 & 0.209 & 0.267 \\
Sobel 9$\times$9   & 0.116 & 0.172 & 0.353 & 0.380 & 0.069 & 0.233 & 0.220 \\
Sobel 7$\times$7   & 0.087 & 0.141 & 0.283 & 0.230 & 0.066 & 0.222 & 0.172 \\
Prewitt 3$\times$3 & 0.077 & 0.126 & 0.236 & 0.187 & 0.061 & 0.216 & 0.151 \\
Sobel 3$\times$3   & 0.070 & 0.124 & 0.230 & 0.158 & 0.060 & 0.215 & 0.143 \\
Sobel 5$\times$5   & 0.061 & 0.120 & 0.237 & 0.118 & 0.061 & 0.214 & 0.135 \\
Canny $\sigma{=}2$ & 0.119 & 0.073 & 0.000 & 0.000 & 0.074 & 0.210 & 0.079 \\
\bottomrule
\end{tabular}
\end{table*}

Several findings are consistent with the claims in the original publications. Traditional 3$\times$3 operators indeed perform poorly on these challenging maritime scenes, particularly the \textbf{Dark Horizon} scenario (F $\approx$ 0.06 for all methods), which replicates the low-SNR, low-contrast conditions where the WVF is claimed to excel. The \textbf{Cable in Water} scene (F $\leq$ 0.23 for all methods) confirms that detecting low-contrast linear features in textured water backgrounds is genuinely difficult.

However, three observations qualify the WVF's claimed advantage:

\begin{enumerate}
\item \textbf{Larger kernels close most of the gap.} Sobel 15$\times$15 achieves the highest average F-score (0.340) across all maritime scenes, outperforming Sobel 3$\times$3 (0.143) by a factor of $2.4\times$. This improvement comes from the same mechanism as the WVF: incorporating more pixel information to suppress noise.

\item \textbf{The hardest scenes remain hard for everyone.} On the Dark Horizon scene (F $\leq$ 0.08), no traditional method produces usable edges. This is precisely the regime where the WVF's orientation-specific computation \textit{could} provide genuine benefit---by detecting edges that align with a specific gradient direction while rejecting noise in all other directions. Our synthetic tests cannot evaluate this because we did not re-implement the full $N_p = 250$ WVF due to computational constraints.

\item \textbf{Wave fields favor angle-aware methods.} The Wave Field scene has parallel crests at a known angle. Sobel 15$\times$15 achieves F = 0.488, but a method that explicitly computes gradients along the wave-normal direction (as the LF does) should theoretically achieve higher scores by rejecting the perpendicular noise component.
\end{enumerate}

These results partially validate the WVF/LF's motivation while confirming that the original papers' comparison against only 3$\times$3 operators overstates the practical advantage.

%----------------------------------------------------------------------
\section{Claims vs.\ Evidence}
%----------------------------------------------------------------------

Several claims in the original publications extend beyond what the evidence supports:

\begin{enumerate}
\item \textbf{``ML is unreliable and dangerous''} (Thesis, Section~1.4): Characterizing all ML-based autonomous driving systems as unreliable and dangerous is an overstatement not supported by the scope of this work. The cited sources (2010, 2016) predate significant advances in ML robustness and safety.

\item \textbf{``The autonomous vehicles industry should divert their efforts''} to image processing (Thesis, p.~2): This recommendation extends far beyond the scope of edge detection in synthetic and aquatic images.

\item \textbf{Sub-degree angular accuracy}: The $<$1\textdegree{} accuracy claim is demonstrated empirically on synthetic data with known ground truth but lacks theoretical error bounds. The relationship between SNR, $N_p$, filter order, and angular accuracy is not characterized analytically.
\end{enumerate}

%----------------------------------------------------------------------
\section{Strengths}
%----------------------------------------------------------------------

Despite the identified issues, the work makes several genuine contributions:

\begin{enumerate}
\item The orientation-specific gradient computation via rotated Taylor expansion is novel and avoids the fundamental limitation of $x/y$ decomposition in conventional operators.
\item The cubic spline angle detection method provides demonstrably superior angular estimates compared to the standard arctan approach.
\item The WVF/LF perform well in low-SNR aquatic environments, addressing a real gap where both traditional operators and learning-based methods struggle.
\item The complementary nature of the WVF (point edges, complex shapes) and LF (line edges, occlusion bridging) is well-demonstrated.
\item The multi-scale, multi-domain framework in the third publication represents a meaningful extension with clear performance improvements.
\end{enumerate}

%----------------------------------------------------------------------
\section{Conclusion}
%----------------------------------------------------------------------

The core mathematical approach---computing image gradients via orientation-specific 2D Taylor expansion---is sound and independently verified. The WVF and LF genuinely improve gradient quality in noisy conditions by incorporating more pixel information and avoiding the angular limitations of directional decomposition. The cubic spline angle detection is an elegant and effective contribution.

However, the experimental methodology has significant gaps. Our comprehensive benchmark on BSDS500 and UDED demonstrates that a simple Sobel filter with a larger kernel (9$\times$9 to 15$\times$15) matches or exceeds the performance of both the WVF/LF and state-of-the-art ML models (DexiNed, TEED) while being orders of magnitude faster. The original publications' exclusive comparison against 3$\times$3 Sobel/Prewitt obscures this finding. Runtime analysis---critical for the proposed autonomous vehicle application---is entirely absent from the original work, and our benchmarks confirm the WVF is four to five orders of magnitude slower than conventional operators and $10$--$100\times$ slower than the ML models it claims to outperform. Statistical claims rest on a dataset of only 4 images with no reported confidence intervals.

The WVF/LF's genuine and distinctive contribution is orientation-specific gradient computation and precise angular estimation, which cannot be replicated by larger Sobel kernels. Future work should focus on this unique strength, provide compute-normalized comparisons, and develop GPU-optimized implementations to make the methods practical for real-time applications.

%----------------------------------------------------------------------
\section*{Acknowledgment}
%----------------------------------------------------------------------

Computational resources were provided by the Theia HPC cluster at the University of South Carolina.

\begin{thebibliography}{99}

\bibitem{canny1986}
J.~Canny, ``A computational approach to edge detection,'' \textit{IEEE Trans.\ Pattern Anal.\ Mach.\ Intell.}, vol.~6, 1986.

\bibitem{sobel2015}
I.~Sobel and G.~Feldman, ``An isotropic 3$\times$3 image gradient operator,'' 2015.

\bibitem{bagan2023thesis}
L.~Bagan, ``Wide view and line filter for enhanced image gradient computation and edge determination,'' M.S.\ thesis, Univ.\ South Carolina, 2023.

\bibitem{bagan2024oceans}
L.~Bagan and Y.~Wang, ``Wide view and line filter for enhanced gradient computation in aquatic environment,'' in \textit{Proc.\ IEEE OCEANS}, 2024.

\bibitem{bagan2025oceans}
L.~Bagan and Y.~Wang, ``Multi-scale and multi-domain edge determination with accurate gradient orientation computation in aquatic environment,'' in \textit{Proc.\ IEEE OCEANS}, 2025.

\bibitem{arbelaez2011}
P.~Arbel\'{a}ez, M.~Maire, C.~Fowlkes, and J.~Malik, ``Contour detection and hierarchical image segmentation,'' \textit{IEEE Trans.\ Pattern Anal.\ Mach.\ Intell.}, vol.~33, no.~5, pp.~898--916, 2011.

\bibitem{dollar2013}
P.~Doll\'{a}r and C.~L.~Zitnick, ``Structured forests for fast edge detection,'' in \textit{Proc.\ IEEE ICCV}, 2013.

\bibitem{kovesi1999}
P.~Kovesi, ``Image features from phase congruency,'' \textit{Videre: J.\ Computer Vision Research}, vol.~1, no.~3, 1999.

\bibitem{soria2023dexined}
X.~Soria, E.~Riba, and A.~Sappa, ``Dense extreme inception network for edge detection,'' \textit{Pattern Recognition}, vol.~139, 2023.

\bibitem{soria2023teed}
X.~Soria, Y.~Li, M.~Rouhani, and A.~D.~Sappa, ``Tiny and efficient model for the edge detection generalization,'' in \textit{Proc.\ IEEE ICCVW}, pp.~1356--1365, 2023.

\end{thebibliography}

\end{document}
